Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-22 LICS21/intro.tex
Remove duplicate references
Introduction: added "(under double-blind review elsewhere)"

Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08}). 
A recent result \cite{MakOPW20} proves that AST programs have density functions that are differentiable almost everywhere.
This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.

The AST problem is not just important, but also difficult: 
Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
In recent unpublished work, the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.

>>>>

Moreover various theorems about probabilistic programs rely on the assumption that the program terminates almost surely (see e.g.~\cite{DBLP:conf/uai/GoodmanMRBT08}). 
A recent result \cite{MakOPW21} proves that AST programs have density functions that are differentiable almost everywhere.
This is significant for Bayesian machine learning, because almost everywhere differentiability is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.

The AST problem is not just important, but also difficult: 
Deciding AST of first-order imperative programs with discrete probabilities is $\Pi^0_2$-complete \cite{DBLP:conf/mfcs/KaminskiK15}.
In recent unpublished work (under double-blind review elsewhere), the AST problem for higher-order programs with continuous distributions (and suitable primitive functions) is shown to have the same complexity.
Though the problem of proving AST of imperative probabilistic programs is much studied \cite{DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05,DBLP:conf/popl/FioritiH15,DBLP:conf/pldi/ChenH20,DBLP:journals/toplas/ChatterjeeFNH18}, to our knowledge the problem of proving AST of higher-order programs with continuous distributions is new.
